Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-05-19 LaTeX/Turnover.tex
one sentence

\title{Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems}
\author{Michael F Scott$*^1$, Matthew M Osmond$*^2$, and Sarah P Otto$^2$}
\date{}
\maketitle
\noindent
$*$ These authors contributed equally to this work

@@

Animals and angiosperms exhibit extremely diverse sex determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, and Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found in the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

Predominant theories in accounting for the spread of new sex determination systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
\citet{vanDoorn:2007eu,vanDoorn:2010hu} show that new sex-determining loci can be favoured if they arise in closer linkage with a locus that experiences sexual antagonism. 
For example, linkage allows favourable associations to build up between a male-beneficial allele and a neo-Y chromosome. 
Such associations can favour a new master sex-determining gene on a new chromosome \citep{vanDoorn:2007eu} and can also favour a transition between male and female heterogamety \citep[e.g., a ZW to XY transition,][]{vanDoorn:2010hu}. 
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determination locus will develop similar, favourable associations and select against the spread of a new sex-determination system. 
Here we extend these studies by explictly calculating the the equilibrium allele frequencies of loci that are very tightly linked to the ancestral sex-determining region.

@@

There are various ways in which a period of haploid selection could influence transitions between sex determination systems. 
If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
However, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

@@

%genetic system
We consider the transition between ancestral and novel sex determination systems using a three locus model. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination (ESD) such that zygotes develop as females in a proportion ($k$) of the environments they (randomly) experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\Hermaphrodite}$, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Any linear order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Individuals that are heterozygous at the \textbf{A} locus may experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection, and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  A & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  a & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  Aa & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  aa & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Tranmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  a & $1-\alpha^\Hermaphrodite=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c, \ref{eq:recursions}d, \ref{eq:recursions}g, \ref{eq:recursions}h). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully epistatically dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the (multiplicative) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination (see table \ref{tab:haplotype_growth}).
The spread of the mutant $m$ allele depends on the frequency of alleles at the other two loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele in female gametes (eggs) from an XX female, $p^\female_X$, and in X-bearing, $p^\male_X$, and Y-bearing, $p^\male_Y$, male gametes (sperm).
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males.

@@

\textcolor{red}{CHECK: DO THE HAPLOID MEAN FITNESSES HAVE TO BE IN THE DENOMINATOR?}

% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

@@

% SOME OF THIS IS REPEATED IN THE APPENDIX CURRENLTY
We are particularly concerned with the conditions under which a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other and invasion depends on the recombination rate between the \textbf{M} and \textbf{A} loci ($R$) see equations \eqref{eq:neoYR} and \eqref{eq:neoWR}.

@@

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $S_{A}=(D^\male +\alpha_{\Delta}^\male+t^\male) - (D^\female+\alpha_{\Delta}^\female+t^\female)$ is the difference in fitness in males versus females for the $A$ allele against the $a$ allele across diploid selection, gametic competition, and meiosis. 
$D^\Hermaphrodite=\big{(} \bar{p}s^\Hermaphrodite+(1-\bar{p})h^\Hermaphrodite s^\Hermaphrodite\big{)} -\big{(} \bar{p}h^\Hermaphrodite s^\Hermaphrodite+(1-\bar{p}) \big{)}$ is the difference in fitness between $A$ and $a$ alleles in diploids of sex $\Hermaphrodite \in \{\female,\male\}$; $\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex (see Appendix).

@@

Equation \eqref{eq:lambda_neoW} shows that if there is no haploid selection ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), as considered by \citet{vanDoorn:2010hu}, the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W',XY}=\lambda_{Y',XY}$) such that heterogametic transitions (XY to ZW or ZW to XY) can also occur only if the neo-sex-determining region is more closely linked to a locus under selection ($R<r$). 
However, if there is any haploid selection, the additional term in equation \eqref{eq:lambda_neoW} can be positive, which can allow, for example, neo-W invasion ($\lambda_{W',XY}>1$) even when the neo-sex-determining region is less closely linked to the selected locus ($R>r$). 
These transitions are unusual because, when $R>r$, associations that have built up between alleles more favourable in one sex and that sex will be weakened. 
Therefore, mean fitness can decrease (Figure \ref{fig:Combination_MeanFit}B,D).

We find that neo-W alleles can invade an XY system for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$ (e.g., in the ancestor the selected locus is on an autosome and the novel sex-determining allele arises on it). 
In table \ref{tab:specialcases} we give the conditions where invasion occurs when we further assume that haploid selection only occurs in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured during diploid selection in males and females ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B). %, which is 50\% of the parameter space. 
When there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D). %, which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W uncouples the sex-determining locus from a locus under selection.

>>>>

\title{Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems}
\author{Michael F Scott$*^1$ and Matthew M Osmond$*^2$, and Sarah P Otto$^2$}
\date{}
\maketitle
\noindent
$*$ These authors contributed equally to this work

@@

Animals and angiosperms exhibit extremely diverse sex determination systems \citep[reviewed in][]{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx}. 
Among species with genetic sex determination of diploid sexes, some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? 
mammals and most dioecious plants \citep{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining region changes. 
For example, transitions of the master sex-determining gene between chromosomes or the evolution of new master sex-determining genes have occurred in Salmonids \citep{Li:2011fm,Yano:2012di}, Diptera \citep{Vicoso:2015hf}, and \textit{Oryzias} \citep{Myosho:2012fv}.
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
In addition, many gonochoric clades with genetic sex determination exhibit transitions between male (XY) and female (ZW) heterogamety, including lizards \citep{Ezaz:2009tk}, eight of 26 teleost fish families \citep{Mank:2006bt}, true fruit flies \citep[Tephritids,][]{Vicoso:2015hf}, amphibians \citep{Hillis:1990gu}, the angiosperm genus \textit{Silene} \citep{Slancarova:2013dq}, Coleoptera and Hemiptera \citep[][plate 2]{Beukeboom:2014vb}.
Indeed, in some cases, both male and female heterogametic sex determination systems can be found in the same species, as exhibited by some cichlid species \citep{Ser:2010iq} and \textit{Rana rugosa} \citep{Ogata:2007jm}.
In addition, multiple transitions have occurred between genetic and environmental sex determination systems, e.g., in reptiles and fishes \citep{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc}.

@@

Predominant theories in accounting for the spread of new sex determination systems by selection involve fitness differences between sexes (e.g., sexually antagonistic selection) or sex ratio selection. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
\citet{vanDoorn:2007eu,vanDoorn:2010hu} show that new sex-determining loci can be favoured if they arise in closer linkage with a locus that experiences sexual antagonism. 
For example, linkage allows favourable associations to build up between a male-beneficial allele and a neo-Y chromosome. 
Such associations can favour a new master sex-determining gene on a new chromosome \citep{vanDoorn:2007eu} and can also favour a transition between male and female heterogamety \citep[e.g., a ZW to XY transition,][]{vanDoorn:2010hu}. 
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determination locus will develop similar, favourable associations and select against the spread of a new sex-determination system. 
Here, we extend these studies by explictly calculating the the equilibrium allele frequencies of loci that are very tightly linked to the ancestral sex-determining region.

@@

There are various ways in which a period of haploid selection could influence transitions between sex determination systems. 
Firstly, if we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
In this respect, we might expect that haploid selection would affect transitions between sex determination systems in a similar manner to sex-specific diploid selection \citep[as explored by][]{vanDoorn:2007eu,vanDoorn:2010hu}. 
That is, new masculizing mutations (neo-Y chromosomes) could be favoured via associations with alleles that are beneficial in the male haploid stage. 
However, sex ratios can also become biased by linkage between the sex-determining region and a locus that harbours genetic variation in haploid fitness. 
For example, there are several known cases of sex ratio bias caused by sex-linked meiotic drive alleles \citep[][, Chapter 3]{Burt:2006} or selection among X- and Y-bearing pollen \citep{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
It is not immediately clear how the spread of new sex determination systems would be influenced by the combination of sex ratio biases and associations between haploid selected loci and sex-determining regions.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model. 
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X}; $XX$ genotypes become females and $XY$ become males (or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is epistatically dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination (ESD) such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination, where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before gametic competition. 
A full description of our model, including recursion equations, is given in the Appendix. 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition) separately. 
Selection during gametic competition depends on the \textbf{A} locus genotype, relative fitnesses are given by $w_A^\Hermaphrodite$ and $w_a^\Hermaphrodite$ ($\Hermaphrodite \in \{\female,\male\}$; see table \ref{tab:fitnesstable}). %$w_{A}^\male$ and $w_{a}^\male$ for male gametes and $w_{A}^\female$ and $w_{a}^\female$ for female gametes, see table \ref{tab:fitnesstable}. 
We assume that all gametes compete for fertilization during gametic competition, which is not the case for monogamous mating systems where gametes from only one mating partner are present. 
Gametic competition in monogamous mating systems is equivalent to meiotic drive in our model, which only alters the frequency of gametes produced by heterozygotes. 
After gametic competition, random mating occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, relative fitnesses are given by $w_{g}^{\male}$ in males and $w_{g}^{\female}$ in females, where $g$ is the diploid genotype at the \textbf{A} locus ($g \in \{AA, Aa, aa\}$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes is then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination (i.e., an odd number of cross-overs) occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus may experience meiotic drive; $Aa$ heterozgotes of sex $\Hermaphrodite$ produce gametes bearing allele $A$ with probability $\alpha^\Hermaphrodite$. 
Thus, the \textbf{A} locus can experience sex-specific gametic competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Relative fitness of different genotypes in sex $\Hermaphrodite \in \{\female,\male\}$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Relative fitness during gametic competition \\ [0.5ex] \hline
  A & $w_{A}^\Hermaphrodite = 1+t^\Hermaphrodite$ \\
  a & $w_{a}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Relative fitness during diploid selection \\ [0.5ex] \hline
  AA & $w_{AA}^\Hermaphrodite = 1+ s^\Hermaphrodite$ \\
  Aa & $w_{Aa}^\Hermaphrodite = 1+h^\Hermaphrodite s^\Hermaphrodite$ \\
  aa & $w_{aa}^\Hermaphrodite = 1$ \\ [0.5ex] \hline
  Genotype & Tranmission during meiosis in $Aa$ heterozygotes \\ [0.5ex] \hline
  A & $\alpha^\Hermaphrodite=1/2+\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  a & $(1-\alpha^\Hermaphrodite)=1/2-\alpha_{\Delta}^{\Hermaphrodite}/2$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

@@

The evolution of a new sex determination system requires that a rare mutant allele, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system of eight equations describing the next generation frequency of eggs and sperm carrying the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h). %, evaluated at the above equilibrium.
%Thus, the system describing the change in frequency of the $m$ allele consists of eight recursion equations. 
This system simplifies substantially in a number of cases of interest. 
Dominant neo-Y (when $k=0$) or neo-W alleles (when $k=1$) are only found in male diploids (neo-Y) or female diploids (neo-W) such that their growth rate ultimately depends only on the change in frequency of $m$-bearing gametes produced by males (for a neo-Y) or by females (for a neo-W). 
Furthermore, if the $m$ allele is fully dominant over the ancestral sex-determining system, phenotypes are not affected by the genotype at the ancestral sex-determining region (\textbf{X} locus). 
Thus, the invasion of rare dominant neo-Y or neo-W alleles is determined by the largest eigenvalue that solves the quadratic characteristic polynomial
$\lambda^2+ b \lambda + c =0$.
%\begin{equation}
%\lambda^2+ b \lambda + c =0
%\label{eq:charpoly_neoY}
%\end{equation}
%
%\noindent
%where $b$ is the average of the growth rates of the two haplotypes that carry the $m$ allele ($mA$ and $ma$), $b=(\lambda_{mA}+\lambda_{ma})/2$, and $c$ also involves the fitness of $m$ alleles when they recombine onto the other \textbf{A} background in a heterozygote, $c=\lambda_{mA}\lambda_{ma}+\rho_{mA}\rho_{ma}$ (see table \ref{tab:haplotype_growth}).
In this case $b= - (\lambda_{mA} + \lambda_{ma})$ and $c = \lambda_{mA} \lambda_{ma} -\rho_{mA} \rho_{ma}$, where $\lambda_{mi}$ is the (discrete time) growth rate of mutant haplotypes on background $i\in\{A,a\}$, accounting for loss due to recombination, and $\rho_{mi}$ is the rate of addition of mutant haplotypes onto background $i\in\{A,a\}$ due to recombination, see table \ref{tab:haplotype_growth}.
The spread of the mutant $m$ allele depends on the frequency of alleles at the other loci in the ancestral population. 
In the ancestral population, it is convenient to follow the frequency of the $A$ allele in female gametes (eggs) from an XX female, $p^\female_X$, and in X-bearing, $p^\male_X$, and Y-bearing, $p^\male_Y$, male gametes (sperm).
We also track the fraction of male gametes that are Y-bearing, $q$, which may deviate from $1/2$ due to meiotic drive in males.

@@

\textcolor{red}{CHECK: DO THE HAPLOID MEAN FITNESSES HAVE TO BE IN THE DENOMINATOR?}
% as there are now two 2s in the lambdas of table 2. perhaps the 2 in the denominator is needed because the frequency of males sums to one as does the frequency of females?}

@@

%%% ADD ZYGOTIC SEX RATIO HERE

We are particularly concerned with the conditions under which a rare neo-sex-determining allele increases in frequency, which occurs when the largest eigenvalue, $\lambda$, is greater than one. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other and invasion will occur when the recombination rate between the \textbf{M} and \textbf{A} loci ($R$) is below a critical value (this critical value can be greater than $1/2$ such that unlinked $m$ alleles invade), see equations \eqref{eq:neoYR} and \eqref{eq:neoWR}.

@@

When $\lambda_{ma}>1$, a neo-W linked to the $a$ allele spreads because it is favoured by selection in females despite the fact that the $a$ allele was stably fixed on the Y, which we might commonly assume to harbour male beneficial alleles. 
This situation can occur when there is a male-biased zygotic sex ratio and/or when the $a$ allele has higher fitness in females (given the

Explain $ma$ and $mA$ hypotheses.

We can also simplify cases where $R>0$ by assuming that selection and drive are weak ($s^\Hermaphrodite$, $t^\Hermaphrodite$, $\alpha_{\Delta}^\Hermaphrodite$ of order $\epsilon$), which implicitly assumes that $R$ is relatively large.

@@

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $S_{A}=(D^\male +\alpha_{\Delta}^\male+t^\male) - (D^\female+\alpha_{\Delta}^\female+t^\female)$ is the difference in fitness in males versus females for the $A$ allele against the $a$ allele across diploid selection, gametic competition, and meiosis. 
$D^\Hermaphrodite=\big{(} \bar{p}s^\Hermaphrodite+(1-\bar{p})h^\Hermaphrodite s^\Hermaphrodite\big{)} -\big{(} \bar{p}h^\Hermaphrodite s^\Hermaphrodite+(1-\bar{p}) \big{)}$ is the difference in fitness between $A$ and $a$ alleles in diploids of sex $\Hermaphrodite \in \{\female,\male\}$; $\bar{p}$ is the leading-order probability of mating with an $A$-bearing gamete from the opposite sex, see appendix.

@@

Equation \eqref{eq:lambda_neoW} shows that if there is no selection upon haploid genotypes ($t^\Hermaphrodite=\alpha^\Hermaphrodite_{\Delta}=0$), as considered by \citet{vanDoorn:2010hu}, the spread of a neo-W is equivalent to the spread of a neo-Y ($\lambda_{W',XY}=\lambda_{Y',XY}$) such that heterogametic transitions (XY to ZW or ZW to XY) can also occur only if the neo-sex-determining region is more closely linked to a locus under selection ($R<r$). 
However, if there is any haploid selection, the additional term in equation \eqref{eq:lambda_neoW} can be positive, which can allow, for example, neo-W invasion ($\lambda_{W',XY}>1$) even when the neo-sex-determining region is less closely linked to the selected locus ($R>r$). 
These transitions are unusual because, when $R>r$, associations that have built up between alleles more favourable in one sex and that sex will be weakened. 
Therefore, mean fitness can decrease (Figure \ref{fig:Combination_MeanFit}B,D).

We find that neo-W alleles can invade an XY system for a large number of selective regimes. 
To clarify the parameter space under which $\lambda_{W',XY}>1$, we consider several special cases. 
Firstly, if the \textbf{A} locus is unlinked to the ancestral sex-determining region ($r=1/2$), a more closely linked neo-W ($R<1/2$) can always invade because $\left( \hat{p}^\male_Y-\hat{p}^\male_X \right)=0$ such that the second term in equation \eqref{eq:lambda_neoW} disappears and invasion depends only on the sign of $(r-R)$. 
Indeed, invasion typically occurs when the neo-W is more closely linked to the selected locus than the ancestral sex-determining region (Figure \ref{fig:Combination_Centimorgans}).
Secondly, we can simplify cases where invasion occurs despite $R>r$ using the special case where $R=1/2$ and $r<1/2$. 
In table \ref{tab:specialcases} we give the conditions where invasion occurs where we further assume that haploid selection only occurs in one sex (e.g., during male meiosis only) and dominance coefficients are equal in the two sexes, $h^\female=h^\male$. 
Where there is no gametic competition and meiotic drive in one sex only, an unlinked neo-W can invade as long as the same allele is favoured during diploid selection in males and females ($s^\female s^\male>0$, see Figure \ref{fig:Combination_Centimorgans}B).%, which is 50\% of the parameter space. 
Where there is no meiotic drive and gametic competition occurs in one sex only, an unlinked neo-W can invade as long as the same allele is favoured in male and female diploid selection and there are sex differences in selection of one type (e.g., $s^\female(s^\male-s^\female)>0$, see Figure \ref{fig:Combination_Centimorgans}C,D). %, which is 25\% of the parameter space. 
These special cases indicate that neo-W invasion can occur for a relatively large fraction of parameter space, even if the neo-W uncouples the sex-determining locus from a locus under selection.
